Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-06-30 manuscript/manuscript_file.Rmd

In this study we present the first national-scale survey of N~2~O concentrations and emissions from lakes, reservoirs, and ponds, hereafter referred to as waterbodies, in the conterminous U.S. (CONUS). A generalized random-tessellation survey design (Stevens and Olsen, 2004) was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005? compare my sample IDs to NLA IDs]** sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week (Figure 1). Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate (See NLA website for all data?). `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks (Figure 1). We modeled the data using a bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 224,916 CONUS waterbodies in the target population **[where did I get 224,916, Roy has 465,897. See technical support document]**. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and \>65% of the waterbodies in every ecoregion were undersaturated (Table 1).

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere (@Mosier_etal_1998) and joins other recent reports that lentic waterbodies can function as an N~2~O sink. @Webb_etal_2019 found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and @Soued_etal_2016 report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020) but kortelainen and Soued are mostly pristine lakes]**. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean N~2~O saturation ratio (ratio of measured to equilibrium concentration) was equal to or greater than 1 in seven of the nine ecoregions **[(Table 1, Figure 2)]**, suggesting that, on average, waterbodies in these ecoregions were supersatured with N~2~O and were therefore functioning as a source of N~2~O. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations that exceed equilibrium in most ecoregions can be attributed to the distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies (Figure 2). Across the CONUS, and within each of the nine major ecoregions, the distribution of N~2~O saturation ratios had median values that were less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources.

@@

Dissolved N~2~O concentration increased with NO~3~ concentration **[Fig. 3]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Knowles_1982, @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015, @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975, @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink.  N~2~O consumption via denitrification likely explains the N~2~O undersaturation observed at low NO~3~ concentrations, whereas denitrification was likely a source of N~2~O at higher NO~3~ concentrations.

@@

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@Carpenter_1983). Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (@Beaulieu_etal_2015, @Beaulieu_etal_2014).

@@

**Not sure where I am going with this paragraph. Maybe scratch it. Mostly included as a placeholder for these references** While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O. Perhaps because previous surveys have included little variation in waterbody size (@Webb_etal_2019, @Kortelainen_etal_2020) or water quality (@Soued_etal_2016).

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^ (Table 1). This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) (SI Figure 1). At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[need table or figure.  See ecoreg_stats$post_med_no3 variable.  maybe a SI table with summaries of all data from all_predictions?]**.

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) had the greatest N~2~O flux in all ecoregions, but flux estimates for this lake size class were also the most uncertain (Fig. 6). Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere (Table 1).  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface (Table 1).  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

At the national scale, we estimate that CONUS waterbodies emit [problem knitting inline code? table1_dat %>% filter(`spatial domain` == "national") %>% select(`N2O flux`) %>% pull()] 1049 (358 - 1581) metric tons N~2~O year^-1^, assuming that the emission rates reported here hold for an entire calendar year. While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ (@epa_2023).  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC (@epa_2023), with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00 metric tons N~2~O year^-1^), stationary combustion (84,000 metric tons N~2~O year^-1^), and manure management (66,000 metric tons N~2~O year^-1^).

Our annual flux estimates assume that emission rate estimates based on samples collected during the summer reflect annual conditions.  The literature contains strong evidence of seasonality in N~2~O emission rates, but the patterns differ among studies.  Some systems have maximum N~2~O emission rates during the winter (@Musenze_etal_2014, @Kortelainen_etal_2020), others in the Fall (@Beaulieu_etal_2014), and yet others during the summer (@Beaulieu_etal_2010).  While our assumption of constant emissions across the year adds uncertainty to the annual flux estimate, it is not clear that the estimate is biased high or low.

@@

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted (@Beaulieu_etal_2014).  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simultaneously mitigate N~2~O emissions.

@@

>>>>

In this study we present the first national-scale survey of N~2~O concentrations and emissions from lakes, reservoirs, and ponds, hereafter referred to as waterbodies, in the conterminous U.S. (CONUS). A generalized random-tessellation survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005? compare my sample IDs to NLA IDs]** sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate (See NLA website for all data?). `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 224,916 CONUS waterbodies in the target population **[where did I get 224,916, Roy has 465,897. See technical support document]**. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and \>65% of the waterbodies in every ecoregion were undersaturated (Table 1).

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and Soued et al. [@Soued_etal_2016] report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020) but kortelainen and Soued are mostly pristine lakes]**. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean N~2~O saturation ratio (ratio of measured to equilibrium concentration) was equal to or greater than 1 in seven of the nine ecoregions **[(Table 1, Fig. 2)]**, suggesting that, on average, waterbodies in these ecoregions were supersatured with N~2~O and were therefore functioning as a source of N~2~O. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations that exceed equilibrium in most ecoregions can be attributed to the distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **[(Fig. 2)]**. Across the CONUS, and within each of the nine major ecoregions, the distribution of N~2~O saturation ratios had median values that were less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources.

@@

Dissolved N~2~O concentration increased with NO~3~ concentration **[Fig. 3]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Knowles_1982; @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015; @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975; @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink.  N~2~O consumption via denitrification likely explains the N~2~O undersaturation observed at low NO~3~ concentrations, whereas denitrification was likely a source of N~2~O at higher NO~3~ concentrations.

@@

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@Carpenter_1983). Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

@@

**Not sure where I am going with this paragraph. Maybe scratch it. Mostly included as a placeholder for these references** While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O. Perhaps because previous surveys have included little variation in waterbody size [@Webb_etal_2019; @Kortelainen_etal_2020] or water quality [@Soued_etal_2016].

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^ (Table 1). This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[need table or figure.  See ecoreg_stats$post_med_no3 variable.  maybe a SI table with summaries of all data from all_predictions?]**.

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) had the greatest N~2~O flux in all ecoregions, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere (Table 1).  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface (Table 1).  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

At the national scale, we estimate that CONUS waterbodies emit `r table1_dat %>% janitor::clean_names() %>% filter(spatial_domain == "national") %>% select(n2o_flux) %>% pull()` metric tons N~2~O year^-1^, assuming that the emission rates reported here hold for an entire calendar year. While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ [@epa_2023].  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC [@epa_2023], with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00 metric tons N~2~O year^-1^), stationary combustion (84,000 metric tons N~2~O year^-1^), and manure management (66,000 metric tons N~2~O year^-1^).

Our annual flux estimates assume that emission rate estimates based on samples collected during the summer reflect annual conditions.  The literature contains strong evidence of seasonality in N~2~O emission rates, but the patterns differ among studies.  Some systems have maximum N~2~O emission rates during the winter [@Musenze_etal_2014; @Kortelainen_etal_2020], others in the Fall [@Beaulieu_etal_2014], and yet others during the summer [@Beaulieu_etal_2010].  While our assumption of constant emissions across the year adds uncertainty to the annual flux estimate, it is not clear that the estimate is biased high or low.

@@

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted [@Beaulieu_etal_2014].  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simultaneously mitigate N~2~O emissions.

@@

<div id="refs"></div>

# Methods

## Survey design

The population of interest (POI) for the 2017 NLA survey [@USEPA_NLA17_2022] was CONUS waterbodies of at least 1 hectare in area, at least 1 meter deep, with at least 0.1 hectare of open water, and a minimum residence time of one week **[(Fig. 1)]**. Inferences from the survey samples were assumed relevant to average conditions during the summer index period between June and end of September 2017. The POI was sampled according to a generalized random-tessellation design with stratification [@Stevens_Olsen_2004], which prescribed a spatially balanced, probabilistic sample with stratification along three factors: surface area (hectares; 5 ordered levels), ecoregion [@Omernik_1987; @Herlihy_etal_2008; @Omernik_Griffith_2014] (9 levels), and US state (excluding AK and HI; 48 levels). The size categories were sampled with unequal probability, such that larger lakes were over-sampled relative to smaller lakes. In total, 1091 waterbodies were sampled; and 95 of those waterbodies were re-visited for a second sample. The revisit samples were used to assess out-of-sample model performance.

## Sampling
Samples were collected from approximately the middle of each waterbody for analysis of algal toxins (microcystins and cylindrospermopsin), benthic macroinvertebrates,  phytoplankton, atrazine, water chemistry, chlorophyll‐a, zooplankton, and dissolved N2O.  Duplicate dissolved N~2~O  samples were collected approximately 0.05 m below the water surface by drawing 105 mL of water into a 140 mL syringe equipped with a stopcock and containing 35 mL of air. The syringes were gently shaken for five minutes and the headspace gas transferred to a 30 mL syringe.  Finally, the gas sample was transferred into a 12 mL pre-evacuated (<50 mTorr) glass vial (Labco, Wales, UK) sealed with a PTFE/Silicone stacked on top of a grey chlorobutyl septa.  The temperatue of the water in the syringe was recorded immediately after the gas was transferred into the vial.  Duplicate 20 mL air samples were collected from approximately 2 m above the water surface and transferred to 12 mL glass vials.  A vertical profile of water temperatue was measured at the same site.

Nitrous oxide partial pressure in the gas samples was measured via gas chromatography (Bruker 450, Billerica, MA, USA), an electron capture detector, and autosampler (CTC Analytics, PAL-xt, Zwingen, Switzerland). A 90:10 mixture of argon and methane was used as the carrier gas. The original dissolved N~2~O concentration was calculated from the measured headspace partial pressue, the temperature specific Bunsen coefficient, and a mass balance for the headspace equilibration system as described in Beaulieu et al. [@Beaulieu_etal_2012]. The equilibrium dissolved gas concentration at the time of sampling was calculated from the measured N~2~O partial pressure in the air samples, the measured barometric pressure, and the water temperature. Dissolved N~2~O concentration is presented as a saturation ratio, defined as the ratio of measured to equilibrium concentration.

Analytical, quality assurance, and field sampling details for all measurements can be found at the 2017 National Lakes Assessment Laboratory Operations Manual[@epa_lom], 2017 National Lakes Assessment Quality Assurance Project Plan [@epa_qapp], and 2017 National Lakes Assessment Field Operations Manual [@epa_fom], respectively.  All data can be accessed online [@epa_data].

## Statistical inference

The survey samples were treated statistically as a randomly selected subset of the complete POI, but with a pattern of "missingess" prescribed by the selection algorithm [@Gelman_etal_2014; @Zachmann_etal_2022]. For inferences, estimates were statistically adjusted to account for the selection mechanism using a Multilevel regression with poststratification (MRP) approach [@Gelman_Little_1997; @Park_etal_2004; @Gelman_etal_2020]. MRP is a model-based approach carried out in two steps. In the first step, the sampled responses of interest (_e.g._, N~2~O) are conditioned on the survey design variables (_e.g._, ecoregion, state, size) in a multilevel regression model. Multilevel models are often recommended for this adjustment because they naturally provide (1) regularized estimates along the design groupings; and (2) estimates for design levels that may be missing from the sample, but are part of the population of interest [@Gelman_Little_1997; @Gelman_etal_2020; @Kennedy_Gelman_2021]. In the second step of MRP, the fitted regression parameters are "poststratified" or re-weighted based on a known or assumed distribution of units (_e.g.,_ US waterbodies) in the POI [@Gelman_Little_1997; @Park_etal_2004; @Gelman_etal_2020; @Kennedy_Gelman_2021; @Zachmann_etal_2022]. In this study, because N~2~O flux was to be calculated as a function of waterbody-level properties (_i.e._, $N_2O_{flux} \propto \text{surface area}$), instead of postratifying parameter estimates across groups of units (_e.g._, ecoregions), predictions were made to each of the 465,897 individual waterbodies in the assumed POI and then scaled up for group level inferences by summarizing the full predictive distributions across the groupings of interest. The fully Bayesian implementation made propagation of uncertainty from the waterbody-level predictions to higher level summaries straightforward.

The final model was used to make probabilistic predictions for (1) dissolved and equilibrium N2O concentrations; (2) the N2O saturation ratio; and (3) the proportion of under-saturated water bodies (_i.e._, saturation ratio < 1). Saturation ratio was calculated as the ratio of dissolved to equilibrium N~2~O. Because dissolved and equilibrium N~2~O were each measured on the same sample units (_i.e._, a single waterbody), their joint distribution was modeled in order to account for potential statistical dependencies due to, for example, common dependencies on geography. The modeled response variable was, therefore, multivariate. Although the mean marginal probabilities estimated from separate models may have provided comparable estimates, the joint model allowing correlated errors was expected to better capture uncertainty and potentially improve out-of-sample predictions, should the variables be conditionally correlated [@Warton_etal_2015; @Poggiato_etal_2021]. The saturation ratio was not modeled or predicted directly, but was estimated _post hoc_ as a "derived" quantity [@Link_Barker_2010; @Gelman_etal_2014; @McElreath_2020] by dividing the posterior predictive distribution (PPD) for dissolved N~2~O by the PPD for equilibrium N~2~O. Similarly, the proportion of under-saturated waterbodies was estimated from this PPD as the proportion of waterbodies with a saturation ratio less than 1. All models were fit using the `brms` [@Burkner_2017] package in `R` [@R_Core_Team] as an interface to Stan [@Stan_Development_Team], a software for fitting fully Bayesian models using Hamiltonian Monte Carlo (HMC).

Fully reproducible code and further results and model workflow details are provided in Modeling **Supplement X**. Though our initial models conditioned the response on just the survey design variables, those models consistently resulted in poorer fits to to the observed N~2~O data compared to models with additional covariates. Though the initial models seemed to work well at replicating the central tendency of the dissolved N~2~O data, simulations of other properties of the observed data, such as the variance, skewness, and kurtosis, were less satisfactory. The relatively poor fit to this aspect of the observed data also resulted in a downward bias in estimates of the observed saturation ratio. However, including a measure of NO~3~ concentration as a covariate on dissolved N~2~O resulted in substantially improved posterior predictive model checks. In addition, including surface temperature and elevation measures as covariates on equilibrium N~2~O resulted in substantially improved fits, as did adding flexibility in the distributional terms for both dissolved and equilibrium N~2~O. Both dissolved and equilibrium N~2~O were allowed to vary according to the survey design variables and environmental covariates in the final model presented here.

Including these helpful covariates, however, led to some complexities in the MRP process. To make inferences to the POI, the covariates needed to be (1) fully observed across the POI or (2) their missingness needed to be modeled. For the lake area and elevation covariates, data was available for all lakes in the POI from previously compiled geospatial databases. However, surface temperature and NO~3~ were not observed for lakes outside of the sample. Therefore, a more complex model was developed to include surface temperature and NO3 as additional responses conditioned on the survey design variables and the fully observed covariates. This approach to inference was akin to Bayesian structural equation modeling approaches [@Merkle_etal_2021; @Merkle_Rosseel_2018]. The dependence structure of the represented by this model may be characterized as:

$$\begin{aligned} 
{p( \boldsymbol{N_2O_{diss}}}\ &|\ Survey,\ Area,\ {\boldsymbol{NO_3}},\ {\boldsymbol{Temp}}) \\ 
{p( \boldsymbol{N_2O_{equi}}}\ &|\ Survey,\ Elev,\ {\boldsymbol{Temp}})\\
{p( \boldsymbol{NO_3}}\ &|\ Survey,\ Area,\ {\boldsymbol{Temp}}) \\ 
{p( \boldsymbol{Temp}}\ &|\ Survey,\ Lat,\ Elev,\ Day)
\end{aligned}$$

Variables in bold text above were treated as partially observed with respect to the POI (_i.e._, observed only in the sample), whereas variables not in bold were fully observed. The partially observed variables, including dissolved and equilibrium N~2~O, NO~3~, and surface temperature, were each modeled conditional on the survey design variables and other partially and/or fully observed covariates. In order to make predictions and propagate uncertainty through the dependency structure, the fitted multivariate model was used to first predict surface temperature in the POI, since it depended only on the fully observed covariates. That predictive distribution was then used alongside the relevant fully observed covariates to predict NO~3~ in the POI. Finally, the predictive distributions for temperature and NO~3~ were used in predicting both N~2~O responses in the POI.
